---
title: "Run a minimal deconvolution simulation"
author: "Sean Maden"
date: "2022-12-08"
output: html_document
---

This vignette shows how to run a minimal deconvolution simulation on example 
datasets.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lute)
```

# Simulation setup

```{r}
num.sim <- 1e3
```

We have set the number of simulations to `r num.sim`. We need to define the core 
deconvolution objects to run the simulation. These are the following 3 list variables:

* `lgv` : Marker signals. These are used to calculate $Z$. We can assign these as
follows:

```{r}
marker1 <- c(1, 0)
marker2 <- c(0, 1)
lgv <- lapply(seq(num.sim),function(ii){list(marker1, marker2)})
```

* `lsv` : Size factors. These are used to transform $Z$ and also to make the $Y$ 
pseudo-bulked sample. To assign these values, use:

```{r}
size1 <- 1
size2 <- 100
lsv <- lapply(seq(num.sim), function(ii){c(size1, size2)})
```

* `lpv` : True prediction values. These are used to make the $Y$ pseudo-bulked sample and
to compare predictions returned from deconvolution. We can vary complementary 
proportions for 2 types using:

```{r}
prop1 <- seq(1e-3, 1, 1e-3)
prop2 <- rev(prop1)
lpv <- lapply(seq(num.sim), function(ii){c(prop1[ii], prop2[ii])})
```

# Run simulations

Now run simulations and store the results as `lres`.

```{r}
lres <- decon_analysis(lgv, lpv, lsv)
```

# Analyze results

We access the plot objects from `lres`. 

```{r}
lgg <- lres$lgg
```

We can view the scatter plots of NNLS-predicted proportions for type 1 (x-axis)
by the root mean squared error (RMSE), grouped on whether the $S$-transform was
first applied to the $Z$ reference, using:

```{r}
lgg$ggpt1
```

We can view the violin plots of RMSE grouped on $S$ transformation status using:

```{r}
lgg$ggvp
```

Finally, we can view the scatterplot of the RMSEs for the non-transformed data 
(x-axis) versus the $S$-tranformed data (y-axis), with a red reference line 
(y-intercept = 0, slope = 1), using:

```{r}
lgg$ggpt2
```